---
title: "red wine quality analysis"
output: html_document
---
Explore Red Wine Quality by Nan Li
========================================================
### Introduction

In this project, we use R and apply exploratory data analysis (EDA) techniques to explore the dataset of wine quality and physicochemical properties. The objective is to explore which chemical properties influence the quality of red wines. And we also produce refined plots to illustrate interesting relationships in the data. The background information of the data is available at this [link](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityReds.csv) and descriptions of data is [here](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt).

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(dplyr)
library(GGally)
library(Hmisc)
library(corrplot)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv('wineQualityReds.csv')
```
### Descriptive statistics 

We'll start with the data structure first. The dataset contains 13 variables and 1599 observations. For each variable, we also have its descriptive statistics for initial observations.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Check the variables
names(wine)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
str(wine)
```
X and quality are discrete variables. All other variables seem to be continuous numerical quantities.
From the variable names and descriptions, it appears that fixed.acidity ~ volatile.acidity and free.sulfur.dioxide ~ total.sulfur.dioxide may have correlations with each other.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine)
```
Since we are primarily interested in quality variable, it would also be interesting to notice the basic statistics on that as well.
*quality* is an ordered, categorical, discrete variable. From the literature, this was on a 0-10 scale, and was rated by at least 3 wine experts. The values ranged only from 3 to 8, with a mean of 5.6 and median of 6.



# Univariate Plots Section

We'll draw quick histograms for these 12 variables and see the pattern for each distributions.

### Histograms
```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots}
grid.arrange(
  ggplot(aes(x=quality), data = wine)+ geom_bar(),
  ggplot(aes(x=fixed.acidity), data = wine) +
  geom_histogram(),
  ggplot(aes(x=volatile.acidity), data = wine) +
  geom_histogram(),
  ggplot(aes(x=citric.acid), data = wine) +
  geom_histogram(),
  ncol=2)


```


```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(
  ggplot(aes(x=residual.sugar), data = wine) +
  geom_histogram(),
  ggplot(aes(x=chlorides), data = wine) +
  geom_histogram(),
  ggplot(aes(x=free.sulfur.dioxide), data = wine) +
  geom_histogram(),
  ggplot(aes(x=total.sulfur.dioxide), data = wine) +
  geom_histogram(),
  ncol=2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(
  ggplot(aes(x=density), data = wine) +
  geom_histogram(),
  ggplot(aes(x=pH), data = wine) +
  geom_histogram(),
  ggplot(aes(x=sulphates), data = wine) +
  geom_histogram(),
  ggplot(aes(x=alcohol), data = wine) +
  geom_histogram(),
  ncol=2
)
```

### Box-Plots
We also draw Boxplots for each variables as another indicator of the distributions. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(
  ggplot(aes(x='',y=wine$fixed.acidity),data = wine, ylab='')+
    geom_boxplot() +labs(x="fixed\nacidity", y=""),

    ggplot(aes(x='',y=wine$volatile.acidity),data = wine, ylab='')+
    geom_boxplot() +labs(x="volatile\nacidity", y=""),
  
   ggplot(aes(x='',y=wine$citric.acid),data = wine, ylab='')+
    geom_boxplot() +labs(x="citric\nacid", y=""),
  
   ggplot(aes(x='',y=residual.sugar),data = wine, ylab='')+
    geom_boxplot() +labs(x="resid.sugar", y=""),

    ggplot(aes(x='',y=chlorides),data = wine, ylab='')+
    geom_boxplot() +labs(x="chlorides", y=""),
  
  ncol=5
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(
 
   ggplot(aes(x='',y=free.sulfur.dioxide),data = wine, ylab='')+
    geom_boxplot() +labs(x="free SO2", y=""),
  
  ggplot(aes(x='',y=total.sulfur.dioxide),data = wine, ylab='')+
    geom_boxplot() +labs(x="total SO2", y=""),
  
    ggplot(aes(x='',y=density),data = wine, ylab='')+
    geom_boxplot() +labs(x="density", y=""),
  
  ggplot(aes(x='',y=pH),data = wine, ylab='')+
    geom_boxplot() +labs(x="pH", y=""),
  
  ggplot(aes(x='',y=sulphates),data = wine, ylab='')+
    geom_boxplot() +labs(x="sulphates", y=""),
  
  ggplot(aes(x='',y=alcohol),data = wine, ylab='')+
    geom_boxplot() +labs(x="alcohol", y=""),
  
  ncol=6
)
```

# Univariate Analysis

### What is the structure of your dataset?
The dataset contains 13 variables and 1599 observations. It appears that density and pH are normally distributed, with few outliers. Fixed and volatile acidity, sulfur dioxides, sulphates, and alcohol seem to be long-tailed.


### What is/are the main feature(s) of interest in your dataset?
The most interesting factor of this dataset is quality. It has a discrete range of 3-8, we can roughly see that there is normal distribution pattern. A large majority of the wines examined received ratings of 5 or 6, and very few received 3, 4, or 8.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
The variables are the physicochemical attributes of red wine, so with basic chemistry background, the concentration of one chemical may have correlation with other relative chemicals or chemicals with similar components or structure. For example, there are three different acidity attributes, and as pH is defined as a numeric scale used to specify the acidity, so pH could be regarded as the characteristic of wine acidity. 

### Did you create any new variables from existing variables in the dataset?
For further investigation, we plan to create new ordered variable for quality, as it will be more convenient to use in the bivariate or multivariate analysis. 

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
From the box-plots charts, we can see that all variables have outliers, and mostly outliers are on the larger side. Residual sugar and chlorides have extreme outliers. Citric acid have a large number of zero values. 
Alcohol has an irregular shaped distribution but it does not have pronounced outliers.

In order to see more details about the distribution trend of each variable, we can adjust the binwidth, choose proper scale or eliminate the outliers to tidy the data for a smoother visualization. 

### Box-plot Statistics 
To adjust the We'll use the statistics of the boxplot as the x-scale range, so that some of the outliers could be eliminated. Finer histograms of each variable are shown below. For variable *Residual_Sugar* and *choloride*, as it is long-tail skewed, we also draw the histogram in log scale base for smoother distribution. 
```{r echo=FALSE, message=FALSE, warning=FALSE}
  # fixed.acidity  
  boxplot.stats(wine$fixed.acidity)$stats
  
  #'volatile.acidity'
  boxplot.stats(wine$volatile.acidity)$stats
  
  #'citric.acid'
  boxplot.stats(wine$citric.acid)$stats

  #'residual.sugar'
  boxplot.stats(wine$residual.sugar)$stats
  
  #'chlorides'
  boxplot.stats(wine$chlorides)$stats
  
  #'free.sulfur.dioxide'
  boxplot.stats(wine$free.sulfur.dioxide)$stats
  
  #'total.sulfur.dioxide'
  boxplot.stats(wine$total.sulfur.dioxide)$stats
  
  #'sulphates'
  boxplot.stats(wine$sulphates)$stats
  
  #'density'
  boxplot.stats(wine$density)$stats
  
  #'pH'
  boxplot.stats(wine$pH)$stats
  
  #'alcohol'
  boxplot.stats(wine$alcohol)$stats


```
### Acidity

```{r echo=FALSE, message=FALSE, warning=FALSE}

grid.arrange(
ggplot(aes(x=fixed.acidity), data = wine) +
  geom_histogram(binwidth = 0.1)+xlim(4.6,12.3),

ggplot(aes(x=volatile.acidity), data = wine) +
  geom_histogram(binwidth = 0.01)+ xlim(0.12,1.01),

ggplot(aes(x=citric.acid), data = wine) +
  geom_histogram(binwidth = 0.01)+xlim(0,0.8),

ncol=1
)
```

### Residual Sugar
```{r echo=FALSE, message=FALSE, warning=FALSE}

grid.arrange(
  ggplot(aes(x=residual.sugar), data = wine) +
  geom_histogram(binwidth = 0.05)+
  scale_x_continuous(breaks = seq(1,4,0.5), limits =c(1,4)),
  
  ggplot(aes(x=residual.sugar), data = wine) +  
    geom_histogram(binwidth = 0.05)+
    scale_x_log10(breaks = seq(1,15,1))+
    xlab('residual.sugar, log10'),
  
  ncol=1
)
```

### Chlorides
```{r echo=FALSE, message=FALSE, warning=FALSE}

grid.arrange(
  ggplot(aes(x=chlorides), data = wine) +
  geom_histogram(binwidth = 0.001)+
  scale_x_continuous(breaks = seq(0.04,0.12,0.01), limits =c(0.04,0.12)),
  
  ggplot(aes(x=chlorides), data = wine) +  
    geom_histogram(binwidth = 0.001)+
    scale_x_log10(breaks=seq(0,0.6,0.1))+
    xlab('chlorides, log10'),
  
  ncol=1
)
```

### Sulfur Dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE}


grid.arrange(
  ggplot(aes(x=free.sulfur.dioxide),data = wine)+
  geom_histogram(binwidth = 1)+
  xlim(0,42),
  
    ggplot(aes(x=total.sulfur.dioxide),data = wine)+
  geom_histogram(binwidth = 1)+
  xlim(6,122),
  
  ncol=1
  )

```

### Density & pH

```{r echo=FALSE, message=FALSE, warning=FALSE}

grid.arrange(
  ggplot(aes(x=density), data = wine) +
  geom_histogram(binwidth = 0.0002)+xlim(0.9923,1.001),
  
  ggplot(aes(x=pH), data = wine) +
  geom_histogram(binwidth = 0.01)+xlim(2.9,3.7),
  ncol=1)
  

```

### Sulphates

```{r echo=FALSE, message=FALSE, warning=FALSE}

grid.arrange(
  ggplot(aes(x=sulphates), data = wine) +
  geom_histogram(binwidth = 0.01) +xlim(0.33,0.99),
  ggplot(aes(x=sulphates), data = wine) +
    geom_histogram(binwidth = 0.02) +
    scale_x_log10(breaks=seq(0.3,2,0.3)) +
    xlab('sulphates, log10'),
  ncol=1)
```

### Alcohol

```{r echo=FALSE, message=FALSE, warning=FALSE}

  ggplot(aes(x=alcohol), data = wine) +
  geom_histogram(binwidth = 0.1) +xlim(8.4,13.5)
 
```

------

# Bivariate Plots Section

In order to investigate the relationship between two variables, we'd start with calculating the correlations between each variable in the database, then pick the pairs of variables with stronger correlations for further analysis.  

```{r echo=FALSE, message=FALSE, warning=FALSE}

my_fn <- function(data, mapping, ...){
  p <- ggplot(data = wine, mapping = mapping) + 
    geom_point(alpha=1/10) + 
    geom_smooth(method=loess, fill="red", color="red", ...) +
    geom_smooth(method=lm, fill="blue", color="blue", ...)
  p
}
g = ggpairs(wine,columns = 2:13, lower = list(continuous = my_fn))
g
```

Below is another correlation plot, the filled circle shows the strength of correlation between two variable, bigger size with darker color indicates stronger correlation, while smaller and brighter circle indicates weaker correlation. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# create new subset without X
wine.sub <- subset(wine, select = -X)
cor.Matric <-cor(wine.sub)
corrplot(cor.Matric,method='circle',is.corr = FALSE,
         tl.col = "black", tl.srt = 45)
```

From the correlations charts, we can see that some correlation in pairs with stronger correlations like:

fixed.acidity vs. citric.acid
fixed.acidity vs. pH
fixed.acidity vs. density
volatile.acidity vs. citric.acid
free.sulfur.dioxide vs total.sulfur.dioxide
chlorides vs. sulphates
alcohol vs. density
quality vs. alcohol

Create new ordered quality variable for later analysis. The original quality variable is in integer format, but the new one is in categorical factor format, so that the dataset will be categorized into six groups with the label quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# create new variable as ordered factor for quality 
wine$quality_factor <- factor(wine$quality, ordered = T)
str(wine$quality_factor)
table(wine$quality_factor)
```

### Acidity 

```{r echo=FALSE, message=FALSE, warning=FALSE}
#acidity correlations
grid.arrange (
  ggplot(data = wine, aes(x = fixed.acidity, y = citric.acid)) +
  geom_point(alpha=1/2),
  
  ggplot(data = wine, aes(x = volatile.acidity, y = citric.acid)) +
  geom_point(alpha=1/2),
  ncol=1)
```

### Acidity vs. pH

```{r echo=FALSE, message=FALSE, warning=FALSE}
#acidity correlations
grid.arrange (
  ggplot(data = wine, aes(x = fixed.acidity, y = pH)) +
  geom_point(alpha=1/2),
  
  ggplot(data = wine, aes(x = citric.acid, y = pH)) +
  geom_point(alpha=1/2),
  ncol=1)
```

### Free SO2 vs. Total SO2

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wine, aes(x = free.sulfur.dioxide, y = total.sulfur.dioxide)) +
  geom_point(alpha=1/2)
```

### Alcohol vs.Density

In this plot, we can see there correlation between volatile.acidity vs.alcohol, the correlation factor is -0.202288. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=alcohol, y=density), data = wine) +
  geom_point(alpha=1/2)+
  geom_smooth(method = 'lm', color='purple2')

cor.test(wine$volatile.acidity, wine$alcohol)

```

Another density/alcohol correlation plot, the y-axis of purple line represent the median density for wine with same alcohol level. Sample statistics are also shown as below. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
wine.density_by_alcohol <- wine %>%
  group_by(alcohol) %>%
  summarise(density_mean=mean(density),
            density_median=median(density),
            n=n()) %>%
  arrange(alcohol)

head(wine.density_by_alcohol)
  
ggplot(aes(x=alcohol, y=density), data = wine) +
  geom_point(alpha=1/2,color='orange')+
  geom_smooth(stat = 'summary', fun.y=median, color='purple3')+
  xlim(8,14)
```

### Sulfate vs. Alcohol

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wine, aes(y = log10(sulphates), x = alcohol)) +
  geom_point(alpha=1/2)+
  geom_smooth( method='lm',color='purple2')
```

We also facet the data points by quality to compare the correlation in different quality level. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = wine, aes(y = log10(sulphates), x = alcohol)) +
  facet_wrap(~quality)+
  geom_point(alpha=1/5,aes(color=quality_factor))+
  scale_fill_brewer(type='div')
```

### Quality vs. Alcohol

```{r echo=FALSE, message=FALSE, warning=FALSE}
#quality vs. alcohol box-plot
ggplot(aes(x=quality_factor,y=alcohol),data = wine)+
  geom_boxplot()

by(wine$alcohol,wine$quality_factor,summary)

```

```{r echo=FALSE, Bivariate_Plots}
ggplot(aes(x=alcohol),data = wine)+
  geom_histogram(aes(color=quality_factor))+
  facet_wrap(~quality)+
  scale_fill_brewer(type='div')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# scatter-plot
ggplot(aes(x=quality,y=alcohol),data = wine)+
  geom_jitter(aes(color=quality_factor), alpha=1/2,size=1)+
  scale_x_continuous(breaks = seq(3,8,1))+
  scale_color_brewer(type='div')
```

### Quality vs. Volatile.acidity

```{r echo=FALSE, message=FALSE, warning=FALSE}
#quality vs. acidity box-plot

ggplot(aes(x=quality_factor,y=volatile.acidity),data = wine)+
  geom_boxplot()

by(wine$volatile.acidity,wine$quality_factor,summary)
```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?
Wine quality is correlated with level of alcohol and volatile acidity. When volatile acidity decreases, the wine quality increases. For alcohol level of wine, wine quality increase as alcohol level increases, but this trend is not dominating for wine of quality 3,4. 


### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
There is strong correlation between acidity variables, pH and acidity, free.sulfur.dioxide and total.sulfur.dioxide. 


### What was the strongest relationship you found?
The strongest correlation is between fixed.acidity and citric.acid, with correlation factor 0.6717.

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor.test(wine$fixed.acidity, wine$citric.acid)
```

------

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}
# fixed.acitity vs. pH by density
ggplot(aes(x=fixed.acidity, y=pH), data = wine)+
  geom_point(aes(color=density))+
  geom_smooth()+
  scale_color_gradient(low = "white", high = "purple3")
```

```{r echo=FALSE, Multivariate_Plots}
ggplot(aes(x=alcohol, y=volatile.acidity), data = wine)+
  geom_point(aes(color=quality_factor))+
  scale_color_brewer(type='div')
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?
As we have examined before in the bivariate section, we know that quality has strong correlation with volatile.acidity and alcohol. In this section, we mapped dotted chart for volatile.acidity vs.alcohol and colored the dots by its quality factor, so that in this plot, we can see there correlation between volatile.acidity vs.alcohol, the correlation factor is -0.202288, which is not a strong correlation. 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One , echo=FALSE, message=FALSE, warning=FALSE}
 ggplot(aes(x=alcohol), data = wine) +
  geom_histogram(binwidth = 0.1, color=I('black'), fill=I('#F79420')) +xlim(8.4,13.5)+
   ggtitle('Histogram of Alcohol Level Distribution ') +
  ylab('Count') +
  xlab('Alcohol (% volume)')
```

### Description One
As alcohol content is an important factor for wine quality, so we choose the histogram as one of the three plots. The distribution of alcohol level is relatively left-skewed and the most frequent alcohol level is 9.5.

### Plot Two
```{r echo=FALSE, Plot_Two}
# scatter-plot
ggplot(aes(x=quality_factor,y=alcohol,fill=quality_factor),data = wine)+
  geom_boxplot() +
  ggtitle('Box-Plot of Alcohol Levels in Different Wine Qualities') +
  xlab('Quality') +
  ylab('Alcohol (% volume)')
```

### Description Two
This boxplot demonstrates the relationship between alcohol content and wine quality. Generally, higher alcohol content correlated with higher wine quality. However, the median alcohol level of wine with lower quality (3,4) are almost the same.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(x=alcohol, y=volatile.acidity), data = wine)+
  geom_point(aes(color=quality_factor))+
  scale_color_brewer(type='div')+
   ggtitle('Alcohol vs. Volatile Acidity in Wine Quality') +
  ylab('Volatile Acidity (g / dm^3)') +
  xlab('Alcohol (% volume)')
```

### Description Three

As the correlation tests show, wine quality was affected most strongly by alcohol and volatile acidity. And we can conclude that better wine would have relative higher alcohol content and lower volatile acidity.

------

# Reflection

Through this exploratory data analysis, we can reach the following conclusions,
- Mostly frequent quality levels of red wine are 5 and 6.
- When alcohol percentage decreases, density grows.
- When fixed acidity increases density increases as well. 
- Acidity variables are strongly correlated with each other. 

According to my investigation I may conclude that the key factors that determine the wine quality are alcohol content and volatile acidity level. 

For future exploration of this data I would pick one category of wine (for example, quality level 3-4, 5-6, 7-8) to look at the patterns which can appear in each of these three buckets. 